DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life

### 2019-09-30 paper_drafts/bioarxiv_2019.09/report.tex
for reproducibility's sake

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{2}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin = 1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{}
    \pretitle{\vspace{\droptitle}}
  \posttitle{}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage{caption}
\usepackage{blindtext}
\usepackage[left]{lineno}
\linenumbers

@@

\textbf{Title}: DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life

\textbf{Authors}: Luna L. Sánchez-Reyes\textsuperscript{1}, Brian C. O'Meara\textsuperscript{1}

\textbf{Correspondence address}:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \emph{Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA}
\end{enumerate}

\textbf{Corresponding authors}: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}, \href{mailto:bomeara@utk.edu}{\nolinkurl{bomeara@utk.edu}}

@@

Clade ages represent a fundamental piece of information for evolutionary
understanding in many areas of research, from developmental to conservation biology (Felsenstein \protect\hyperlink{ref-Felsenstein1985a}{1985}; Webb \protect\hyperlink{ref-Webb2000}{2000}), from historical biogeography to species diversification studies (Posadas et al. \protect\hyperlink{ref-posadas2006historical}{2006}; Morlon \protect\hyperlink{ref-Morlon2014}{2014}). The primary information needed for these time estimates comes from the fossil record. Coupled with phylogenies with branch lengths based on molecular and/or morphological data, the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
The number of studies publishing phylogenies with branch lengths proportional to geological time (hereafter chronograms) have constantly increased in number for the last two decades (Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}).
Still, generating a chronogram is not an easy task unless you have specialized training: it requires inferring a tree, understanding what fossil data are available and their limits, and where fossils go on the tree. That is why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information (Webb and Donoghue \protect\hyperlink{ref-webb2005phylomatic}{2005}; Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013}).

Wide interest from the scientific community to make information from phylogenies in general and chronograms in particular available for consultation and reuse has spurred the creation of public platforms with various goals and characteristics. TreeBASE (Morell \protect\hyperlink{ref-morell1996roots}{1996}; Piel et al. \protect\hyperlink{ref-Piel2002}{2002}), the Dryad repository (\url{http://datadryad.org/}), and the Open Tree of Life (OToL; Hinchliff et al. \protect\hyperlink{ref-Hinchliff2015}{2015}) platforms store and make available published phylogenies and chronograms for easy scientific reuse. All of them can be queried using automatised web procedures, which permit personalized, large scale queries that are also very fast.
OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates). Treebase and Dryad repositories also contain trees from all groups of life, but the former did not store branch length information until recently (and lacks consistent metadata on what any branchlengths stored mean) and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.

Besides keeping a repository to easily store and share expert phylogenetic and chronogram knowledge, OToL also has the primary goal of synthesising all trees in their repository to expose to the community a single tree of all life depicting the phylogenetic relationships among known lineages.
All or parts of this synthetic tree can be reused for any purpose. However, it currently only focus on synthesizing tree topology, meaning that it does not expose branch length data of any type.
The Timetree of Life project focuses on the synthesis of a single chronogram of life (Hedges et al. \protect\hyperlink{ref-Hedges2006}{2006}) and presents a very accessible, attractive interface. However, the thousands of chronograms this NSF-funded project have compiled for synthesis are only publicly available for visual examination in their website or for download as images, but large scale download remains prohibited by their site. The latest version of their synthetic chronogram (Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}) can be queried only through their website in a non-automatised fashion, and only subsets of it can be reused for analyses with the permission of the authors.
Other platforms such as SuperSmart (Antonelli et al. \protect\hyperlink{ref-antonelli2017supersmart}{2017}) and phylogenerator (Pearse and Purvis \protect\hyperlink{ref-pearse2013phylogenerator}{2013}) are focused in automatised \emph{de novo} chronogram inference, by reusing DNA sequence data to reconstruct phylogenetic trees. However, expert fossil information necessary for subsequent molecular dating analyses still needs to be compiled and curated by the user, rendering them a challenging tool to obtain data on time of lineage divergence for the non-specialist. Moreover, these tools do not provide information from already created expert chronograms.

A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fully public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al. \protect\hyperlink{ref-Stoltzfus2013}{2013}).
In here we present the formal description and implementation of the \texttt{datelife} service, constituted by an R package and a web site (\url{http://www.datelife.org/}). There is still much room for improvement, and flaws and limitations are addressed below. We strived for the current implementation of \texttt{datelife} to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.

@@

To gather, process, and present information, \texttt{datelife} builds up from functions
available in several R packages including rotl (Michonneau et al. \protect\hyperlink{ref-Michonneau2016}{2016}), ape (Paradis et al. \protect\hyperlink{ref-Paradis2004}{2004}),
geiger (Harmon et al. \protect\hyperlink{ref-Harmon2008}{2008}), paleotree (Bapst \protect\hyperlink{ref-Bapst2012a}{2012}), bold (Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phytools (Revell \protect\hyperlink{ref-Revell2012}{2012}), taxize (Chamberlain and Szöcs \protect\hyperlink{ref-Chamberlain2013}{2013}; Chamberlain \protect\hyperlink{ref-Chamberlain2018}{2018}), phyloch (Heibl \protect\hyperlink{ref-Heibl2008}{2008}), phylocomr (Ooms and Chamberlain \protect\hyperlink{ref-Ooms2018}{2018}) and rphylotastic (O'Meara et al. \protect\hyperlink{ref-Omeara2019}{2019}).

A \texttt{datelife} search currently accepts scientific names only. It can be any named clade or binomial specific.
Chronogram search is performed at the species level, so when input names correspond
to named clades, \texttt{datelife} pulls all accepted species names within the
clade from OToL's reference taxonomy to perform the search.
Searches at the infraspecies level are not currently allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
\texttt{datelife} processes input names with the taxon name resolution service (TNRS; Boyle et al. \protect\hyperlink{ref-Boyle2013}{2013}),
which corrects potentially misspelled names and typos, and standardizes spelling
variations and synonyms , increasing the probability of correctly finding the
queried taxa in \texttt{datelife}'s chronogram database.

@@

Information from source chronograms can be summarised with a summary statistic of tree branch lengths, such as median or mean. A much slower, but possibly more accurate Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths (Criscuolo et al. \protect\hyperlink{ref-Criscuolo2006}{2006}) is also implemented via the ape package (Paradis et al. \protect\hyperlink{ref-Paradis2004}{2004}).
The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree. However, we noticed that the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife\_examples package). Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree. The Branch Length Adjuster (BLADJ) algorithm (Webb et al. \protect\hyperlink{ref-Webb2008}{2008}) was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.

For tree dating, the congruification algorithm described by Eastman et al. (\protect\hyperlink{ref-Eastman2013}{2013})
is implemented to find shared nodes between trees (congruent nodes). The ages of these nodes are then used as calibrations to date any given tree. Currently implemented methods for tree dating are BLADJ, MrBayes (Huelsenbeck and Ronquist \protect\hyperlink{ref-Huelsenbeck2001}{2001}; Ronquist and Huelsenbeck \protect\hyperlink{ref-Ronquist2003}{2003}) and PATHd8 (Britton et al. \protect\hyperlink{ref-Britton2007}{2007}), a non-clock, rate-smoothing dating method.

@@

\texttt{datelife}'s code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor.
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing increases roughly linearly with number of input taxon names, and
increases considerably if TNRS service is activated. Up to ten thousand names can be processed and searched in less than 30 minutes. A name search through the chronogram database with an already processed query can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:runtime1}).
\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}) that were run locally
--using the devtools package (R Core Team \protect\hyperlink{ref-RCoreTeam2018}{2018}), and on a public server --via
GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
present, unit tests cover more than 50\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).

@@

In this section we demonstrate the types of outputs that can be obtained with \texttt{datelife}, using as an example the bird family Fringillidae of true finches. We performed a higher-taxon search to obtain all data on lineage divergence available in \texttt{datelife}'s database for all recognised species within the Fringillidae (475 spp. according to the Open Tree of Life taxonomy). There are 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Fig. \ref{fig:schronograms1}).
Data from these source chronograms was used to generate two types of summary chronograms, median and SDM. As explained in the \texttt{Description}, data from source chronograms was first summarised into a single distance matrix (using either the median or the SDM method) and then the available node ages were used as calibrations points over a consensus tree topology, to obtain a dated tree with the program BLADJ (Fig. \ref{fig:summaries}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM. In both cases, ages are generally consistent with source ages.
Different source chronograms often show substantial variation in ages for clades (see, for example, the ongoing debate about crown group age of angiosperms (Ramshaw et al. \protect\hyperlink{ref-ramshaw1972time}{1972}; Sanderson and Doyle \protect\hyperlink{ref-sanderson2001sources}{2001}; Magallón et al. \protect\hyperlink{ref-magallon2015metacalibrated}{2015}; Barba-Montoya et al. \protect\hyperlink{ref-barba2018constraining}{2018}). For some studies, especially ones based on branch lengths (studies of diversification, timing of events, trait evolution, and more), using a different chronogram may return different results (Title and Rabosky \protect\hyperlink{ref-title2016macrophylogenies}{2016}). Stitching together these chronograms can create a larger tree and uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses still requires more research.

@@

At the moment, \texttt{datelife}'s chronogram database is not very large, storing 231 chronograms up to the time the manuscript was written. This represents 5.79\% of the largest existing chronogram database, which is not open for scientific reuse nor automatised data mining (Kumar et al. \protect\hyperlink{ref-Kumar2017}{2017}). OToL is the only public tree repository from where \texttt{datelife} can currently pull chronograms to construct its database.
A previous version of TimeTree's synthetic chronogram (Hedges et al. \protect\hyperlink{ref-Hedges2015}{2015}) was made available in the OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
This ensures that some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
Thus, incorporation of more published chronograms deposited in OpenTree or perhaps pulled from Dryad directly, to \texttt{datelife}'s database is crucial to improve its services.
Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would still need to be determined by hand.
Consequently, we would like to emphasize on the importance of sharing chronogram data for the scientific community, in repositories that require expert input and manual curation, such as OToL's tree repository.

Another potential concern comes from summary chronograms. We currently summarize all source chronogram data by default. Users can subset source data if they have reasons to favor some or one source chronogram over others. Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely. To our knowledge, there is no objective way to determine if an expert chronogram is better than other. Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used. Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations. It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses (Schenk \protect\hyperlink{ref-schenk2016sec}{2016}). Yet, there are different ways to use secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods). This remains to be tested.

>>>>

\documentclass[11pt,]{article}
\usepackage[left=1in,top=1in,right=1in,bottom=1in]{geometry}
\newcommand*{\authorfont}{\fontfamily{phv}\selectfont}
\usepackage[]{mathpazo}

  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}

\usepackage{abstract}
\renewcommand{\abstractname}{}    % clear the title
\renewcommand{\absnamepos}{empty} % originally center

\renewenvironment{abstract}
 {{%
    \setlength{\leftmargin}{0mm}
    \setlength{\rightmargin}{\leftmargin}%
  }%
  \relax}
 {\endlist}

\makeatletter
\def\@maketitle{%
  \newpage
%  \null
%  \vskip 2em%
%  \begin{center}%
  \let \footnote \thanks
    {\fontsize{18}{20}\selectfont\raggedright  \setlength{\parindent}{0pt} \@title \par}%
}
%\fi
\makeatother

\setcounter{secnumdepth}{0}

\title{DateLife: Leveraging databases and analytical tools to reveal the dated
Tree of Life \thanks{Replication files are available on the author's Github account
(\url{http://github.com/LunaSare}). \textbf{Current version}: September
29, 2019; \textbf{Corresponding authors}:
\href{mailto:lsanche7@utk.edu}{\nolinkurl{lsanche7@utk.edu}};
\href{mailto:bomeara@utk.edu}{\nolinkurl{bomeara@utk.edu}}; 1.
Corresponding adrdress: \emph{Department of Ecology and Evolutionary
Biology, University of Tennessee, Knoxville, 425 Hesler Biology
Building, Knoxville, TN 37996, USA}}  }

\author{\Large Luna L. Sanchez Reyes\vspace{0.05in} \newline\normalsize\emph{University of Tennessee, Knoxville}   \and \Large Brian O'Meara\vspace{0.05in} \newline\normalsize\emph{University of Tennessee, Knoxville}  }

\date{}

\usepackage{titlesec}

\titleformat*{\section}{\normalsize\bfseries}
\titleformat*{\subsection}{\normalsize\itshape}
\titleformat*{\subsubsection}{\normalsize\itshape}
\titleformat*{\paragraph}{\normalsize\itshape}
\titleformat*{\subparagraph}{\normalsize\itshape}

\usepackage{natbib}
\bibliographystyle{apsr}
\usepackage[strings]{underscore} % protect underscores in most circumstances

\newtheorem{hypothesis}{Hypothesis}
\usepackage{setspace}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{hyperref}
\usepackage[left]{lineno}
\linenumbers
\usepackage{caption}

% move the hyperref stuff down here, after header-includes, to allow for - \usepackage{hyperref}

\makeatletter
\@ifpackageloaded{hyperref}{}{%
\ifxetex
  \PassOptionsToPackage{hyphens}{url}\usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \PassOptionsToPackage{hyphens}{url}\usepackage[draft,unicode=true]{hyperref}
\fi
}

\@ifpackageloaded{color}{
    \PassOptionsToPackage{usenames,dvipsnames}{color}
}{%
    \usepackage[usenames,dvipsnames]{color}
}
\makeatother
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Luna L. Sanchez Reyes (University of Tennessee, Knoxville) and Brian O'Meara (University of Tennessee, Knoxville)},
             pdfkeywords = {Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science;
Congruification; Supertree; Calibrations},  
            pdftitle={DateLife: Leveraging databases and analytical tools to reveal the dated
Tree of Life},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

% Add an option for endnotes. -----

% add tightlist ----------
\providecommand{\tightlist}{%
\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% add some other packages ----------

% \usepackage{multicol}
% This should regulate where figures float
% See: https://tex.stackexchange.com/questions/2275/keeping-tables-figures-close-to-where-they-are-mentioned
\usepackage[section]{placeins}

@@

% \pagenumbering{arabic}% resets `page` counter to 1 
%
% \maketitle

{% \usefont{T1}{pnc}{m}{n}
\setlength{\parindent}{0pt}
\thispagestyle{plain}
{\fontsize{18}{20}\selectfont\raggedright 
\maketitle  % title \par

}

{
   \vskip 13.5pt\relax \normalsize\fontsize{11}{12} 
\textbf{\authorfont Luna L. Sanchez Reyes} \hskip 15pt \emph{\small University of Tennessee, Knoxville}   \par \textbf{\authorfont Brian O'Meara} \hskip 15pt \emph{\small University of Tennessee, Knoxville}

}

}

\begin{abstract}

    \hbox{\vrule height .2pt width 39.14pc}

    \vskip 8.5pt % \small

\noindent The combination of new analytical techniques, availability of more
fossil and molecular data, and better practices in data sharing has
resulted in a steady accumulation of chronograms in public and open
databases such as TreeBASE, Dryad, and Open Tree of Life for a large
quantity and diversity of organisms in the last few decades. However,
getting a tree with branch lengths proportional to time remains
difficult for many biologists and the non-academic community, despite
its importance in many areas of research, education, and science
communication. \texttt{datelife} is a service implemented via an R
package and a web site (\url{http://www.datelife.org/}) for efficient
reuse, summary and reanalysis of published data on lineage divergence
times. The main workflow starts with at least two taxon names as input,
either as tip labels on a tree, or as a simple comma separated character
string. A name search is then performed across the chronogram database
and positively identified source trees are pruned to maintain queried
taxa only and stored as a named list of patristic distance matrices.
Source chronogram data can be summarised using branch length summary
statistics or variance minimizing approaches to generate a single
summary chronogram. Source chronogram data can also be used as
calibration points to date a tree containing some or all names from the
initial query. If there is no information available for any queried
taxa, data can be simulated. All source and summary chronograms can be
saved in formats that permit easy reuse and reanalysis. Summary and
newly generated trees are potentially useful to evaluate evolutionary
hypothesis in different areas of research in biology. How well this
trees work for this purpose still needs to be tested. \texttt{datelife}
will be useful to increase awereness on the existing variation in expert
time of divergence data, and might foster exploration of the effect of
alternative divergence time hypothesis on the results of analyses,
nurturing a culture of more cautious interpretation of evolutionary
results.

\vskip 8.5pt \noindent \emph{Keywords}: Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science;
Congruification; Supertree; Calibrations \par

    \hbox{\vrule height .2pt width 39.14pc}

\end{abstract}

\vskip -8.5pt

 % removetitleabstract

\noindent \doublespacing

@@

Clade ages represent a fundamental piece of information for evolutionary
understanding in many areas of research, from developmental to
conservation biology \citep{Felsenstein1985a, Webb2000}, from historical
biogeography to species diversification studies
\citep{posadas2006historical, Morlon2014}. The primary information
needed for these time estimates comes from the fossil record. Coupled
with phylogenies with branch lengths based on molecular and/or
morphological data, the time of divergence of extant and extinct
lineages can be reconstructed with molecular dating methods. The number
of studies publishing phylogenies with branch lengths proportional to
geological time (hereafter chronograms) have constantly increased in
number for the last two decades \citep{Kumar2017}. Still, generating a
chronogram is not an easy task unless you have specialized training: it
requires inferring a tree, understanding what fossil data are available
and their limits, and where fossils go on the tree. That is why there
has been an urge for promoting and facilitating reuse of the vast amount
of phylogenetic and time of lineage divergence data that has been
generated and made available in publications, for the advantage of
research relying on this information
\citep{webb2005phylomatic, Stoltzfus2013}.

Wide interest from the scientific community to make information from
phylogenies in general and chronograms in particular available for
consultation and reuse has spurred the creation of public platforms with
various goals and characteristics. TreeBASE
\citep{morell1996roots, Piel2002}, the Dryad repository
(\url{http://datadryad.org/}), and the Open Tree of Life
\citep[OToL;][]{Hinchliff2015} platforms store and make available
published phylogenies and chronograms for easy scientific reuse. All of
them can be queried using automatised web procedures, which permit
personalized, large scale queries that are also very fast. OToL stores
trees with branch length information from a wide range of living
organisms, implementing a metadata structure that stores the branch
length units (i.e., time or relative susbtitution rates). Treebase and
Dryad repositories also contain trees from all groups of life, but the
former did not store branch length information until recently (and lacks
consistent metadata on what any branchlengths stored mean) and Dryad
stores many other types of biological data using metadata that does not
allow automatic distinction of types of trees and branch length units,
impairing the automatised access to time of lineage divergence
information.

Besides keeping a repository to easily store and share expert
phylogenetic and chronogram knowledge, OToL also has the primary goal of
synthesising all trees in their repository to expose to the community a
single tree of all life depicting the phylogenetic relationships among
known lineages. All or parts of this synthetic tree can be reused for
any purpose. However, it currently only focus on synthesizing tree
topology, meaning that it does not expose branch length data of any
type. The Timetree of Life project focuses on the synthesis of a single
chronogram of life \citep{Hedges2006} and presents a very accessible,
attractive interface. However, the thousands of chronograms this
NSF-funded project have compiled for synthesis are only publicly
available for visual examination in their website or for download as
images, but large scale download remains prohibited by their site. The
latest version of their synthetic chronogram \citep{Kumar2017} can be
queried only through their website in a non-automatised fashion, and
only subsets of it can be reused for analyses with the permission of the
authors. Other platforms such as SuperSmart
\citep{antonelli2017supersmart} and phylogenerator
\citep{pearse2013phylogenerator} are focused in automatised \emph{de
novo} chronogram inference, by reusing DNA sequence data to reconstruct
phylogenetic trees. However, expert fossil information necessary for
subsequent molecular dating analyses still needs to be compiled and
curated by the user, rendering them a challenging tool to obtain data on
time of lineage divergence for the non-specialist. Moreover, these tools
do not provide information from already created expert chronograms.

A tool for efficient reuse of expert, published data on time of lineage
divergence should have an open and fuly public chronogram database
storing data in a format suitable for scientific reuse, an automatised
way of accessing the information, and straightforward means of comparing
and summarizing chronogram information as needed by the user. A
prototype service aiming to meet this characteristics was developed over
a series of hackathons at the National Evolutionary Synthesis Center
\citep{Stoltzfus2013}. In here we present the formal description and
implementation of the \texttt{datelife} service, constituted by an R
package and a web site (\url{http://www.datelife.org/}). There is still
much room for improvement, and flaws and limitations are addressed
below. We strived for the current implementation of \texttt{datelife} to
perform the basic tasks described above, featuring a system for
maintenance of an open database of chronograms pulled from public
repositories, methods to summarize and compare source chronograms, and
new functions to visualize and graphically compare source and summary
chronograms.

@@

To gather, process, and present information, \texttt{datelife} builds up
from functions available in several R packages including rotl
\citep{Michonneau2016}, ape \citep{Paradis2004}, geiger
\citep{Harmon2008}, paleotree \citep{Bapst2012a}, bold
\citep{Chamberlain2018}, phytools \citep{Revell2012}, taxize
\citep{Chamberlain2013, Chamberlain2018}, phyloch \citep{Heibl2008},
phylocomr \citep{Ooms2018} and rphylotastic \citep{Omeara2019}.

A \texttt{datelife} search currently accepts scientific names only. It
can be any named clade or binomial specific. Chronogram search is
performed at the species level, so when input names correspond to named
clades, \texttt{datelife} pulls all accepted species names within the
clade from OToL's reference taxonomy to perform the search. Searches at
the infraspecies level are not currently allowed, so input names
belonging to subspecies or any other infraspecific category are
collapsed to the species level. \texttt{datelife} processes input names
with the taxon name resolution service \citep[TNRS;][]{Boyle2013}, which
corrects potentially misspelled names and typos, and standardizes
spelling variations and synonyms , increasing the probability of
correctly finding the queried taxa in \texttt{datelife}'s chronogram
database.

@@

Information from source chronograms can be summarised with a summary
statistic of tree branch lengths, such as median or mean. A much slower,
but possibly more accurate Super Distance Matrix (SDM) approach for
supertree reconstruction with branch lengths \citep{Criscuolo2006} is
also implemented via the ape package \citep{Paradis2004}. The resulting
summary patristic distance matrix could be clustered with classic
algorithms to return a tree. However, we noticed that the resulting
trees are often non-ultrameric and do not reflect the source chronogram
data (see datelife\_examples package). Instead, we obtained a
distribution of age data from the summary matrix available for nodes on
a consensus tree. The Branch Length Adjuster (BLADJ) algorithm
\citep{Webb2008} was then used to distribute node ages evenly over the
consensus tree, minimizing age variance in the resulting chronogram.

For tree dating, the congruification algorithm described by
\citet{Eastman2013} is implemented to find shared nodes between trees
(congruent nodes). The ages of these nodes are then used as calibrations
to date any given tree. Currently implemented methods for tree dating
are BLADJ, MrBayes \citep{Huelsenbeck2001, Ronquist2003} and PATHd8
\citep{Britton2007}, a non-clock, rate-smoothing dating method.

@@

\texttt{datelife}'s code speed was tested on an Apple iMac with one 3.4
GHz Intel Core i5 processor. We registered variation in computing time
of query processing and search through the database relative to number
of queried taxon names. Query processing increases roughly linearly with
number of input taxon names, and increases considerably if TNRS service
is activated. Up to ten thousand names can be processed and searched in
less than 30 minutes. A name search through the chronogram database with
an already processed query can be performed in less than a minute, even
with a very large number of taxon names (Fig. \ref{fig:runtime1}).
\texttt{datelife}'s code performance was evaluated with a set of unit
tests designed and implemented with the R package testthat
\citep{RCoreTeam2018} that were run locally --using the devtools package
\citep{RCoreTeam2018}, and on a public server --via GitHub, using the
continuous integration tool Travis CI (\url{https://travis-ci.org}). At
present, unit tests cover more than 50\% of \texttt{datelife}'s code
(\url{https://codecov.io/gh/phylotastic/datelife}).

@@

In this section we demonstrate the types of outputs that can be obtained
with \texttt{datelife}, using as an example the bird family Fringillidae
of true finches. We performed a higher-taxon search to obtain all data
on lineage divergence available in \texttt{datelife}'s database for all
recognised species within the Fringillidae (475 spp. according to the
Open Tree of Life taxonomy). There are 13 chronograms containing at
least two Fringillidae species, published in 9 different studies (Fig.
\ref{fig:schronograms1}). Data from these source chronograms was used to
generate two types of summary chronograms, median and SDM. As explained
in the \texttt{Description}, data from source chronograms was first
summarised into a single distance matrix (using either the median or the
SDM method) and then the available node ages were used as calibrations
points over a consensus tree topology, to obtain a dated tree with the
program BLADJ (Fig. \ref{fig:summaries}). Median summary chronograms are
older and have wider variation in maximum ages than chronograms obtained
with SDM. In both cases, ages are generally consistent with source ages.
Different source chronograms often show substantial variation in ages
for clades (see, for example, the ongoing debate about crown group age
of angiosperms
\citep{barba2018constraining, magallon2015metacalibrated, sanderson2001sources, ramshaw1972time}.
For some studies, especially ones based on branch lengths (studies of
diversification, timing of events, trait evolution, and more), using a
different chronogram may return different results
\citep{title2016macrophylogenies}. Stitching together these chronograms
can create a larger tree and uses information from multiple studies, but
the effect of uncertainties and errors here on downstream analyses still
requires more research.

@@

At the moment, \texttt{datelife}'s chronogram database is not very
large, storing 231 chronograms up to the time the manuscript was
written. This represents 5.79\% of the largest existing chronogram
database, which is not open for scientific reuse nor automatised data
mining \citep{Kumar2017}. OToL is the only public tree repository from
where \texttt{datelife} can currently pull chronograms to construct its
database. A previous version of TimeTree's synthetic chronogram
\citep{Hedges2015} was made available in the OToL repository, hence the
amount of lineages represented in datelife's database is at least as
substantial as TimeTree's. This ensures that some information will be
available for any given query, but it does not ensure that the full
state of knowledge of time of divergence data will be available for any
given lineage. Thus, incorporation of more published chronograms
deposited in OpenTree or perhaps pulled from Dryad directly, to
\texttt{datelife}'s database is crucial to improve its services. Methods
to automatically mine chronogram data from the Dryad repository could be
designed and implemented. However, the unit of branch lengths would
still need to be determined by hand. Consequently, we would like to
emphasize on the importance of sharing chronogram data for the
scientific community, in repositories that require expert input and
manual curation, such as OToL's tree repository.

Another potential concern comes from summary chronograms. We currently
summarize all source chronogram data by default. Users can subset source
data if they have reasons to favor some or one source chronogram over
others. Strictly speaking, a good chronogram should reflect the real
time of lineage divergence accurately and precisely. To our knowledge,
there is no objective way to determine if an expert chronogram is better
than other. Some criteria that have been put forward are the level of
lineage sampling and the number of calibrations used. Scientists usually
also favor chronograms coming from studies with primary calibrations to
ones from secondary calibrations. It has been observed with simulations
that divergence times inferred with secondary calibrations are
significantly younger than those inferred with primary calibrations in
analyses performed with bayesian inference methods when priors are
implemented in similar ways in both analyses \citep{schenk2016sec}. Yet,
there are different ways to use secondary calibrations and the bias
might not be encountered with other dating methods that do not require
prior assumptions (such as ML methods). This remains to be tested.

Furthermore, even chronograms obtained with primary data can be very
different, as observed from the comparison of source chronograms in the
Fringillidae example. A large discrepancy in time of lineage divergence
across expert knowledge is well known for different groups of organisms
\citep[e.g., angiosperms;][]{magallon2015metacalibrated}. Comparison of
available chronogram data for a wide range of organisms shown here
suggest that this is a widespread phenomenon that requires further
attention. Characteristics of the data used for dating analyses as well
as from the output chronogram itself, such as quality of alignment
(missing data, GC content), lineage sampling strategy and proportion,
phylogenetic and dating inference method, number of fossils used as
calibrations, support for nodes and ages, and confidence intervals could
be used to score quality of source chronograms. To facilitate subsetting
of source chronograms following different criteria, this information
should be included as metadata manually entered by curators (as is done
in OToL) in the future. Still, even if all source chronograms have been
generated by excellent standards and using similar methods, the
evolutionary history they depict might be very different. Hence,
summarizing chronograms might imply summarizing evolutionary hypothesis.
This could be good from certain point of view, since it could help to
get a single global evolutionary history for a lineage. But it could
also be bad, since we could be loosing part of the evolutionary history
that is only being reflected in some chronograms and not from the
summary chronogram. Ideally, we should still rely on time of lineage
divergence data obtained from a single analysis using fossil data as
primary sources of calibrations, and using fossils that have already
been curated as calibrations to date other trees, which should reflect a
more homogeneous evolutionary history \citep{antonelli2017supersmart}.
This will be implemented in future \texttt{datelife} versions.
